A Dynamic Intrafamily Model of Child Behavior Problems and Birth Timing

### 2018-04-13 inst/paper/diagram_estimates.Rnw
Many fixes

\begin{tikzpicture}[node distance = 2cm, auto]
    % Place nodes
     \node [cloud] (F) {$Shared\ Component_{i}$ (FE)};
     \node [block, above left= 1cm and -1.5cm of F] (T1) {$Mother Age_{1i}$ (T1)};
     \node [block, below left= 1cm and -1.5cm of F] (T2) {$Mother Age_{2i}$ (T2)};
     \node [block, right of=T1, node distance=6cm] (BP1) {$Difficulty_{1i}$ (BP1)};
     \node [block, right of=T2, node distance=6cm] (BP2) {$Difficulty_{2i}$ (BP2)};
     \node [right of=BP1, node distance=3cm] (E1) {\Sexpr{diff_1_err}};
     \node [right of=BP2, node distance=3cm] (E2) {\Sexpr{diff_2_err}};
     \node [left of=T1, node distance=2cm] (pt1) {\Sexpr{m_age_1_err}};
     \node [left of=T2, node distance=2cm] (pt2) {\Sexpr{m_age_2_err}};
     \node [below of=pt1, node distance=1.85cm] (pf) {\Sexpr{fe_err}};
     \node [below right= 2.5cm and 5.3cm of F] (ee) {$ $};
     \node [below right= 2.5cm and -2.5cm of F] (eek) {$ $};
     \node [below = .72cm of E2] (eej) {$ $};
     \node [below right= 0cm and 5.8cm of F, align=center] (ee2) {\Sexpr{b_inst}\\ (\Sexpr{se_inst})};
     \node [left of=F, node distance=5.4cm, align=center] (stt) {\Sexpr{cov_m_age}\\ (\Sexpr{se_cov_m_age})};
     \node [above left= .2cm and 1.4cm of F, align=center] (st1f) {\Sexpr{cov_m_age_1_fe}\\ (\Sexpr{se_cov_m_age_1_fe})};
     \node [below left= .2cm and 1.4cm of F, align=center] (st2f) {\Sexpr{cov_m_age_2_fe}\\ (\Sexpr{se_cov_m_age_2_fe})};
     \node [above right= .7cm and -.1cm of F] (sfbp1) {$1$};
     \node [below right= .2cm and -.1cm of F] (sfbp2) {$1$};
     \node [above right= 1.5cm and -1.9cm of F, align=center] (stbp1) {\Sexpr{b_m_age_1_diff_1}\\ (\Sexpr{se_b_m_age_1_diff_1})};
     \node [below right= .6cm and -1.9cm of F, align=center] (stbp1) {\Sexpr{b_m_age_2_diff_2}\\ (\Sexpr{se_b_m_age_2_diff_2})};
    % Draw edges
    \draw [->] (T1) to (BP1);
    \draw [->] (T2) to (BP2);
    \draw [->] (F) to (BP1);
    \draw [->] (F) to (BP2);
    \draw [->] (E1) to (BP1);
    \draw [->] (E2) to (BP2);
    \draw [->] (pt1) to (T1);
    \draw [->] (pt2) to (T2);
    \draw [->] (pf) to (F);
    \path (pt1) edge [<->, bend right=90, looseness=1.5] node {$ $} (pt2);
    \path (pt1) edge [<->, out=180,in=180] node {$ $} (pf);
    \path (pt2) edge [<->, out=180,in=180] node {$ $} (pf);
    \draw [-] (E1) to [out=0,in=90] (ee2);
    \draw [->] (ee2) to [out=270,in=0] (eej)
    to [out=180, in=0] (eek)
    to [out=180,in=270, looseness=1] (T2);
\end{tikzpicture}

>>>>

\begin{tikzpicture}[node distance = 2cm, auto]
    % Place nodes
     \node [cloud] (F) {$Shared\ Component_{i}$ ($F_{i}$)};
     \node [block, above left= 1cm and -1.5cm of F] (T1) {$Mother Age_{1i}$ ($T_{1i}$)};
     \node [block, below left= 1cm and -1.5cm of F] (T2) {$Mother Age_{2i}$ ($T_{2i}$)};
     \node [block, right of=T1, node distance=6cm] (D1) {$Difficulty_{1i}$ ($D_{1i}$)};
     \node [block, right of=T2, node distance=6cm] (D2) {$Difficulty_{2i}$ ($D_{2i}$)};
     \node [right of=D1, node distance=3cm] (E1) {\Sexpr{diff_1_err}};
     \node [right of=D2, node distance=3cm] (E2) {\Sexpr{diff_2_err}};
     \node [left of=T1, node distance=2cm] (pt1) {\Sexpr{m_age_1_err}};
     \node [left of=T2, node distance=2cm] (pt2) {\Sexpr{m_age_2_err}};
     \node [below of=pt1, node distance=1.85cm] (pf) {\Sexpr{fe_err}};
     \node [below right= 2.5cm and 5.3cm of F] (ee) {$ $};
     \node [below right= 2.5cm and -2.5cm of F] (eek) {$ $};
     \node [below right= 0cm and 5.8cm of F, align=center] (ee2) {\Sexpr{b_inst}\\ (\Sexpr{se_inst})};
     \node [left of=F, node distance=5.4cm, align=center] (stt) {\Sexpr{cov_m_age}\\ (\Sexpr{se_cov_m_age})};
     \node [above left= .2cm and 1.4cm of F, align=center] (st1f) {\Sexpr{cov_m_age_1_fe}\\ (\Sexpr{se_cov_m_age_1_fe})};
     \node [below left= .2cm and 1.4cm of F, align=center] (st2f) {\Sexpr{cov_m_age_2_fe}\\ (\Sexpr{se_cov_m_age_2_fe})};
     \node [above right= .7cm and -.1cm of F] (sfbp1) {$1$};
     \node [below right= .2cm and -.1cm of F] (sfbp2) {$1$};
     \node [above right= 1.5cm and -1.9cm of F, align=center] (stbp1) {\Sexpr{b_m_age_1_diff_1}\\ (\Sexpr{se_b_m_age_1_diff_1})};
     \node [below right= .6cm and -1.9cm of F, align=center] (stbp1) {\Sexpr{b_m_age_2_diff_2}\\ (\Sexpr{se_b_m_age_2_diff_2})};
    % Draw edges
    \draw [->] (T1) to (D1);
    \draw [->] (T2) to (D2);
    \draw [->] (F) to (D1);
    \draw [->] (F) to (D2);
    \draw [->] (E1) to (D1);
    \draw [->] (E2) to (D2);
    \draw [->] (pt1) to (T1);
    \draw [->] (pt2) to (T2);
    \draw [->] (pf) to (F);
    \path (pt1) edge [<->, bend right=90, looseness=1.5] node {$ $} (pt2);
    \path (pt1) edge [<->, out=180,in=180] node {$ $} (pf);
    \path (pt2) edge [<->, out=180,in=180] node {$ $} (pf);
    \draw [->] (E1) to [out=0,in=0] (ee)
    to [out=180,in=0, looseness=.35] (eek)
    to [out=180,in=270, looseness=1] (T2);
\end{tikzpicture}

### 2018-04-13 inst/paper/diagram_estimates.tex
Many fixes

\begin{tikzpicture}[node distance = 2cm, auto]
    % Place nodes
     \node [cloud] (F) {$Shared\ Component_{i}$ (FE)};
     \node [block, above left= 1cm and -1.5cm of F] (T1) {$Mother Age_{1i}$ (T1)};
     \node [block, below left= 1cm and -1.5cm of F] (T2) {$Mother Age_{2i}$ (T2)};
     \node [block, right of=T1, node distance=6cm] (BP1) {$Difficulty_{1i}$ (BP1)};
     \node [block, right of=T2, node distance=6cm] (BP2) {$Difficulty_{2i}$ (BP2)};
     \node [right of=BP1, node distance=3cm] (E1) {35.5};
     \node [right of=BP2, node distance=3cm] (E2) {46.93};
     \node [left of=T1, node distance=2cm] (pt1) {13.8};
     \node [left of=T2, node distance=2cm] (pt2) {15.83};
     \node [below of=pt1, node distance=1.85cm] (pf) {23.23};
     \node [below right= 2.5cm and 5.3cm of F] (ee) {$ $};
     \node [below right= 2.5cm and -2.5cm of F] (eek) {$ $};
     \node [below = .72cm of E2] (eej) {$ $};
     \node [below right= 0cm and 5.8cm of F, align=center] (ee2) {-5.98\\ (2.68)};
     \node [left of=F, node distance=5.4cm, align=center] (stt) {11.94\\ (0.5)};
     \node [above left= .2cm and 1.4cm of F, align=center] (st1f) {13.91\\ (6.23)};
     \node [below left= .2cm and 1.4cm of F, align=center] (st2f) {18.26\\ (8.08)};
     \node [above right= .7cm and -.1cm of F] (sfbp1) {$1$};
     \node [below right= .2cm and -.1cm of F] (sfbp2) {$1$};
     \node [above right= 1.5cm and -1.9cm of F, align=center] (stbp1) {-1.74\\ (0.48)};
     \node [below right= .6cm and -1.9cm of F, align=center] (stbp1) {-1.74\\ (0.48)};
    % Draw edges
    \draw [->] (T1) to (BP1);
    \draw [->] (T2) to (BP2);
    \draw [->] (F) to (BP1);
    \draw [->] (F) to (BP2);
    \draw [->] (E1) to (BP1);
    \draw [->] (E2) to (BP2);
    \draw [->] (pt1) to (T1);
    \draw [->] (pt2) to (T2);
    \draw [->] (pf) to (F);
    \path (pt1) edge [<->, bend right=90, looseness=1.5] node {$ $} (pt2);
    \path (pt1) edge [<->, out=180,in=180] node {$ $} (pf);
    \path (pt2) edge [<->, out=180,in=180] node {$ $} (pf);
    \draw [-] (E1) to [out=0,in=90] (ee2);
    \draw [->] (ee2) to [out=270,in=0] (eej)
    to [out=180, in=0] (eek)
    to [out=180,in=270, looseness=1] (T2);
\end{tikzpicture}

>>>>

\begin{tikzpicture}[node distance = 2cm, auto]
    % Place nodes
     \node [cloud] (F) {$Shared\ Component_{i}$ ($F_{i}$)};
     \node [block, above left= 1cm and -1.5cm of F] (T1) {$Mother Age_{1i}$ ($T_{1i}$)};
     \node [block, below left= 1cm and -1.5cm of F] (T2) {$Mother Age_{2i}$ ($T_{2i}$)};
     \node [block, right of=T1, node distance=6cm] (D1) {$Difficulty_{1i}$ ($D_{1i}$)};
     \node [block, right of=T2, node distance=6cm] (D2) {$Difficulty_{2i}$ ($D_{2i}$)};
     \node [right of=D1, node distance=3cm] (E1) {35.5};
     \node [right of=D2, node distance=3cm] (E2) {46.93};
     \node [left of=T1, node distance=2cm] (pt1) {13.8};
     \node [left of=T2, node distance=2cm] (pt2) {15.83};
     \node [below of=pt1, node distance=1.85cm] (pf) {23.23};
     \node [below right= 2.5cm and 5.3cm of F] (ee) {$ $};
     \node [below right= 2.5cm and -2.5cm of F] (eek) {$ $};
     \node [below right= 0cm and 5.8cm of F, align=center] (ee2) {-5.98\\ (2.68)};
     \node [left of=F, node distance=5.4cm, align=center] (stt) {11.94\\ (0.5)};
     \node [above left= .2cm and 1.4cm of F, align=center] (st1f) {13.91\\ (6.23)};
     \node [below left= .2cm and 1.4cm of F, align=center] (st2f) {18.26\\ (8.08)};
     \node [above right= .7cm and -.1cm of F] (sfbp1) {$1$};
     \node [below right= .2cm and -.1cm of F] (sfbp2) {$1$};
     \node [above right= 1.5cm and -1.9cm of F, align=center] (stbp1) {-1.74\\ (0.48)};
     \node [below right= .6cm and -1.9cm of F, align=center] (stbp1) {-1.74\\ (0.48)};
    % Draw edges
    \draw [->] (T1) to (D1);
    \draw [->] (T2) to (D2);
    \draw [->] (F) to (D1);
    \draw [->] (F) to (D2);
    \draw [->] (E1) to (D1);
    \draw [->] (E2) to (D2);
    \draw [->] (pt1) to (T1);
    \draw [->] (pt2) to (T2);
    \draw [->] (pf) to (F);
    \path (pt1) edge [<->, bend right=90, looseness=1.5] node {$ $} (pt2);
    \path (pt1) edge [<->, out=180,in=180] node {$ $} (pf);
    \path (pt2) edge [<->, out=180,in=180] node {$ $} (pf);
    \draw [->] (E1) to [out=0,in=0] (ee)
    to [out=180,in=0, looseness=.35] (eek)
    to [out=180,in=270, looseness=1] (T2);
\end{tikzpicture}

### 2018-04-13 inst/paper/diagram_model.Rnw
Many fixes

\begin{tikzpicture}[node distance = 2cm, auto]
    % Place nodes
     \node [cloud] (F) {$Shared\ Component_{i}$ (FE)};
     \node [block, above left= 1cm and -1.5cm of F] (T1) {$Mother Age_{1i}$ (T1)};
     \node [block, below left= 1cm and -1.5cm of F] (T2) {$Mother Age_{2i}$ (T2)};
     \node [block, right of=T1, node distance=6cm] (BP1) {$Difficulty_{1i}$ (BP1)};
     \node [block, right of=T2, node distance=6cm] (BP2) {$Difficulty_{2i}$ (BP2)};
     \node [right of=BP1, node distance=3cm] (E1) {$\epsilon_{1i}$};
     \node [right of=BP2, node distance=3cm] (E2) {$\epsilon_{2i}$};
     \node [below right= 2.5cm and 5.3cm of F] (ee) {$ $};
     \node [below right= 2.5cm and -2.5cm of F] (eek) {$ $};
     \node [below right= 0cm and 5.8cm of F] (ee2) {$\sigma_{\epsilon_{1i}T_{2i}}$};
     \node [left of=F, node distance=3.9cm] (stt) {$\sigma_{T_{1i}T_{2i}}$};
     \node [above left= .33cm and .1cm of F] (st1f) {$\sigma_{T_{1i}FE_{i}}$};
     \node [below left= .33cm and .1cm of F] (st2f) {$\sigma_{T_{2i}FE_{i}}$};
     \node [above right= .29cm and .4cm of F] (sfbp1) {$\gamma_{FE_{i}BP_{1i}} = 1$};
     \node [below right= .3cm and .4cm of F] (sfbp2) {$\gamma_{FE_{i}BP_{2i}} = 1$};
     \node [above right= 1.5cm and -1.4cm of F] (stbp1) {$\gamma_{T_{1i}BP_{1i}}$};
     \node [below right= 1.55cm and -1.4cm of F] (stbp2) {$\gamma_{T_{2i}BP_{2i}}$};
    % Draw edges
    \path [line] (T1) -- (BP1);
    \path [line] (T2) -- (BP2);
    \path [line] (F) -- (BP1);
    \path [line] (F) -- (BP2);
    \path [line] (E1) -- (BP1);
    \path [line] (E2) -- (BP2);
    \path (T1) edge [<->, bend right=90, looseness=1.5] node {$ $} (T2);
    \path (T1) edge [<->, out=180,in=180] node {$ $} (F);
    \path (T2) edge [<->, out=180,in=180] node {$ $} (F);
    \draw [->] (E1) to [out=0,in=0] (ee)
    to [out=180,in=0, looseness=.35] (eek)
    to [out=180,in=270, looseness=1] (T2);
\end{tikzpicture}

>>>>

\begin{tikzpicture}[node distance = 2cm, auto]
    % Place nodes
     \node [cloud] (F) {$Shared\ Component_{i}$ ($F_{i}$)};
     \node [block, above left= 1cm and -1.5cm of F] (T1) {$Mother Age_{1i}$ ($T_{1i}$)};
     \node [block, below left= 1cm and -1.5cm of F] (T2) {$Mother Age_{2i}$ ($T_{2i}$)};
     \node [block, right of=T1, node distance=6cm] (D1) {$Difficulty_{1i}$ ($D_{1i}$)};
     \node [block, right of=T2, node distance=6cm] (D2) {$Difficulty_{2i}$ ($D_{2i}$)};
     \node [right of=D1, node distance=3cm] (E1) {$\epsilon_{1i}$};
     \node [right of=D2, node distance=3cm] (E2) {$\epsilon_{2i}$};
     \node [below right= 2.5cm and 5.3cm of F] (ee) {$ $};
     \node [below right= 2.5cm and -2.5cm of F] (eek) {$ $};
     \node [below right= 0cm and 5.8cm of F] (ee2) {$\sigma_{\epsilon_{1i}T_{2i}}$};
     \node [left of=F, node distance=3.9cm] (stt) {$\sigma_{T_{1i}T_{2i}}$};
     \node [above left= .33cm and .1cm of F] (st1f) {$\sigma_{T_{1i}F_{i}}$};
     \node [below left= .33cm and .1cm of F] (st2f) {$\sigma_{T_{2i}F_{i}}$};
     \node [above right= .29cm and .4cm of F] (sfD1) {$\gamma_{F_{i}D_{1i}} = 1$};
     \node [below right= .3cm and .4cm of F] (sfD2) {$\gamma_{F_{i}D_{2i}} = 1$};
     \node [above right= 1.5cm and -1.4cm of F] (stD1) {$\gamma_{T_{1i}D_{1i}}$};
     \node [below right= 1.55cm and -1.4cm of F] (stD2) {$\gamma_{T_{2i}D_{2i}}$};
    % Draw edges
    \draw [->] (T1) to (D1);
    \draw [->] (T2) to (D2);
    \draw [->] (F) to (D1);
    \draw [->] (F) to (D2);
    \draw [->] (E1) to (D1);
    \draw [->] (E2) to (D2);
    \path (T1) edge [<->, bend right=90, looseness=1.5] node {$ $} (T2);
    \path (T1) edge [<->, out=180,in=180] node {$ $} (F);
    \path (T2) edge [<->, out=180,in=180] node {$ $} (F);
    \draw [->] (E1) to [out=0,in=0] (ee)
    to [out=180,in=0, looseness=.35] (eek)
    to [out=180,in=270, looseness=1] (T2);
\end{tikzpicture}

### 2018-04-13 inst/paper/paper.Rmd
Many fixes

Note: The below material is pre-existing and does not account for the covariates noted in the new introduction above, but I included it in case you are interested in the preliminary work.

@@

This study uses the National Longitudinal Survey of Youth 1979 (NLYS79), a probability sample of 12,686 Americans between the ages of 14 and 21 during 1979. Beginning in 1986, female respondents who had children responded to interviews about their children and parenting behavior and assessments were conducted on the children resulting in their addition to a child and young adult sample of the NLYS79. As of 2012, the child and young adult sample consists of 11,512 children, which is estimated to be some 95% of expected childbirths for original NLSY79 participants; as nearly all births to this cohort should be accounted for, avoiding issues with selection on birth timing. Because the focus of this work is the relationship between perceived early-childhood behavior problems and birth timing, analyses are limited to a subsample of mothers with two or more children for which data on infant behavior problems were collected. These perceptions of early child behavior problems were collected from mothers of children between ages 0 and 23 months. This subsample contains 1531 out of 4932 (31%) mothers and 5386 of the 11,512 children (47%). The omitted mothers either had only one child (1699) or had only one child after data collection began in 1986 (1702). This sample is further reduced to 1392 mothers due to missingness in responses to components of the outcome variables.

The outcome variables in this analysis are the Difficulty Composite Raw Scores for both siblings which consists of the sum of ordinal (1 to 5) responses to 11 temperament questions capturing a range of child behaviors, such as “How often do you have trouble soothing or calming your infant when he/she is crying or upset?” and “During the average day, how often does your infant get fussy and irritable?” This scale has an observed range of 11 to 54 (out of possible 11 to 55) and is treated as a continuous measure as it is approximately normally distributed, though there is evidence of departures from normality for difficulty of the second child (see appendix). The predictors are maternal age at birth for each child. Mothers in the analysis subsample were 19 to 40 years of age at birth of first child and 21 to 42 for the second child.

```{r diagrammodel, eval = TRUE, warning=FALSE, message=FALSE, out.width='80%', fig.cap="Conceptual Model"}
# This is a ridiculous hack: I convert the PDF to a PNG on the fly and display it
invisible(pdftools::pdf_convert("diagram_model.pdf", filenames="diagram_model.png", dpi = 300, verbose=FALSE))
knitr::include_graphics("diagram_model.png")
```

Figure \@ref(fig:diagrammodel) depicts the conceptual model of mother’s age at birth and perceived child difficulty. T1 and T2 represent mother’s age at birth of each child which are assumed to be correlated with one another and also with unobserved time-invariant family and mother characteristics (or the shared component), FE. T1, T2, and FE predict child difficulty (BP1 and BP2), and the effect of the shared component is constrained to unity for both outcomes, yielding a sibling fixed effects model. Lastly, σ(ε1,T2) represents the covariance between the disturbance term for the perceived difficulty of the first child (ε1) and mother’s age at second birth (T2). Modeling this covariance accounts for the possibility that child difficulty will impact future fertility decisions and thus timing of future births. We hypothesize that 1) perceived difficulty of the first child is associated with delays in subsequent births and 2) a negative relationship between mother’s age and child difficulty will remain in the presence of any feedback effect of difficulty on second births.

@@

```{r diagramestimates, eval = TRUE, warning=FALSE, message=FALSE, out.width='80%', fig.cap="Model Estimates"}
invisible(pdftools::pdf_convert("diagram_estimates.pdf", filenames="diagram_estimates.png", dpi = 300, verbose=FALSE))
knitr::include_graphics("diagram_estimates.png")
```

Figure \@ref(fig:diagramestimates)

```{r resultsdata, include=FALSE}
load("../data/derived_data/naive_tidy.RData")

```

Figure 2 depicts maximum likelihood estimates of the specified structural equations obtained with LISREL 9.2. A strong negative relationship is observed between mother’s age at birth of child and perceived difficulty in early childhood. Net of time-invariant mother and family characteristics, each additional year older at childbirth corresponds to almost two points lower on the difficulty composite or an approximately 1:1 relationship in standardized metrics (B*=  `r round(naive_tidy$std.all[3],2)` and `r round(naive_tidy$std.all[4], 2)`). Additionally, we see a statistically significant (t=-2.275) relationship between the unaccounted for variation in first child difficulty and mother’s age at second birth, though this relationship is unexpectedly negative. Greater difficulty in early childhood, net of that associated with maternal age, appears associated with modestly earlier second births. As the model is just identified, tests of overall model fit cannot be conducted. For comparison the BP1 error to T2 covariance was restricted to zero to provide fit statistics, though this restricted model is clearly misspecified given the significance of the parameter in the full model. In this model, the root mean square error of approximation is 0.0583 which indicates moderately good fit, though this may be inflated by the presence of a single degree of freedom. The restricted model chi-square is 5.731 (p=0.017), indicating worse fit than the full (saturated) model.

@@

These results indicate that child behavior problems may influence later fertility decisions, rendering estimates of the effect of maternal age biased in standard sibling models, however the effect of maternal age remains in the presence of this feedback. This both provides support for the findings in @hao_family_2006 regarding teen parenting and draws attention to the importance of accounting for dynamic effects in sibling models as described by both @rosenzweig_sisters_1995 and @phillips_sibship_1999. Curiously, the feedback effect of child behavior problems appears to be in a direction counter to expectation; positive variation in child difficulty is associated with a shorter interval to second birth. Extension of this analysis to include relevant covariates from in Hao and Matsueda may shed light on this finding.

@@

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

@@

##### pagebreak

>>>>

Note: The below material is pre-existing and does not account for the covariates noted above, but I included it in case you are interested in the preliminary work.

@@

```{r resultsdata, include=FALSE}
load("../data/derived_data/naive_tidy.RData")
load("../data/derived_data/child_difficulty.RData")
```

This study uses the National Longitudinal Survey of Youth 1979 (NLYS79), a probability sample of 12,686 Americans between the ages of 14 and 21 during 1979. Beginning in 1986, female respondents who had children responded to interviews about their children and parenting behavior and assessments were conducted on the children resulting in their addition to a child and young adult sample of the NLYS79. As of 2012, the child and young adult sample consists of 11,512 children, which is estimated to be some 95% of expected childbirths for original NLSY79 participants; as nearly all births to this cohort should be accounted for, avoiding issues with selection on birth timing. Because the focus of this work is the relationship between perceived early-childhood behavior problems and birth timing, analyses are limited to a subsample of mothers with two or more children for which data on infant behavior problems were collected. These perceptions of early child behavior problems were collected from mothers of children between ages 0 and 23 months. This subsample contains 1531 out of 4932 (31%) mothers and 5386 of the 11,512 children (47%). The omitted mothers either had only one child (1699) or had only one child after data collection began in 1986 (1702). This sample is further reduced to `r nrow(child_difficulty)` mothers due to missingness in responses to components of the outcome variables.

The outcome variables in this analysis are the Difficulty Composite Raw Scores for both siblings which consists of the sum of ordinal (1 to 5) responses to 11 temperament questions capturing a range of child behaviors, such as “How often do you have trouble soothing or calming your infant when he/she is crying or upset?” and “During the average day, how often does your infant get fussy and irritable?” This scale has an observed range of `r min(child_difficulty$Difficulty_1)` to `r max(child_difficulty$Difficulty_1)` (out of possible 11 to 55) for first children and `r min(child_difficulty$Difficulty_2)` to `r max(child_difficulty$Difficulty_2)` for second children. The scale is treated as a continuous measure as it is approximately normally distributed, though there is evidence of mild departures from normality for difficulty of the second child (see appendix). The predictors are maternal age at birth for each child. Mothers in the analysis subsample were `r min(child_difficulty$m_age_at_birth_1)` to `r max(child_difficulty$m_age_at_birth_1)` years of age at birth of first child and `r min(child_difficulty$m_age_at_birth_2)` to `r max(child_difficulty$m_age_at_birth_2)` for the second child
<br/><br/>
```{r diagrammodel, eval = TRUE, warning=FALSE, message=FALSE, out.width='80%', fig.cap="Conceptual Model"}
# This is a ridiculous hack: I convert the PDF to a PNG on the fly and display it
invisible(pdftools::pdf_convert("diagram_model.pdf", filenames="diagram_model.png", dpi = 500, verbose=FALSE))
knitr::include_graphics("diagram_model.png")
```

Figure \@ref(fig:diagrammodel) depicts the conceptual model of mother’s age at birth and perceived child difficulty. $T_{1}$ and $T_{2}$ represent mother’s age at birth of each child which are assumed to be correlated with one another and also with unobserved time-invariant family and mother characteristics (or the shared component), $F_{i}$. $T_{1i}$, $T_{2i}$, and $F_{i}$ predict child difficulty ($D_{1i}$ and $D_{2i}$), and the effect of the shared component is constrained to unity for both outcomes, yielding a sibling fixed effects model. Lastly, $\sigma_{\epsilon_{1i},T_{2i}}$ represents the covariance between the disturbance term for the perceived difficulty of the first child ($\epsilon_{1i}$) and mother’s age at second birth ($T_{2i}$). Modeling this covariance accounts for the possibility that child difficulty will impact future fertility decisions and thus timing of future births. We hypothesize that 1) perceived difficulty of the first child is associated with delays in subsequent births and 2) a negative relationship between mother’s age and child difficulty will remain in the presence of any feedback effect of difficulty on second births.

@@

```{r diagramestimates, eval = TRUE, warning=FALSE, message=FALSE, out.width='80%', fig.cap="Model Estimates"}
invisible(pdftools::pdf_convert("diagram_estimates.pdf", filenames="diagram_estimates.png", dpi = 500, verbose=FALSE))
knitr::include_graphics("diagram_estimates.png")
```

Figure \@ref(fig:diagramestimates) depicts maximum likelihood estimates of the specified structural equations obtained with LISREL 9.2. A strong negative relationship is observed between mother’s age at birth of child and perceived difficulty in early childhood. Net of time-invariant mother and family characteristics, each additional year older at childbirth corresponds to almost two points lower on the difficulty composite or an approximately 1:1 relationship in standardized metrics ($B\*$=  `r round(naive_tidy$std.all[3],2)` and `r round(naive_tidy$std.all[4], 2)`). Additionally, we see a statistically significant (t=-2.275) relationship between the unaccounted for variation in first child difficulty and mother’s age at second birth, though this relationship is unexpectedly negative. Greater difficulty in early childhood, net of that associated with maternal age, appears associated with modestly earlier second births. As the model is just identified, tests of overall model fit cannot be conducted. For comparison the $D_{1}$ error to $T_{2}$ covariance was restricted to zero to provide fit statistics, though this restricted model is clearly misspecified given the significance of the parameter in the full model. In this model, the root mean square error of approximation is 0.0583 which indicates moderately good fit, though this may be inflated by the presence of a single degree of freedom. The restricted model chi-square is 5.731 (p=0.017), indicating worse fit than the full (saturated) model.

@@

These results indicate that child behavior problems may influence later fertility decisions, rendering estimates of the effect of maternal age biased in standard sibling models, however the effect of maternal age remains in the presence of this feedback. This both provides support for the findings in @hao_family_2006 regarding teen parenting and draws attention to the importance of accounting for dynamic effects in sibling models as described by both @rosenzweig_sisters_1995 and @phillips_sibship_1999. Curiously, the feedback effect of child behavior problems appears to be in a direction counter to expectation; positive variation in child difficulty is associated with a shorter interval to second birth. Extension of this analysis to include relevant covariates from @hao_family_2006 may shed light on this finding.

@@

@@
